```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.align="center")
```

## Maximum Likelihood Estimation - Advanced

#### Statistics for Laboratory Scientists ( 140.615 )

### Example from class: Gaussian data

Simulate some Gaussian data.

```{r}
set.seed(1)
n <- 100
x <- rnorm(n,mean=10,sd=4)
```

The maximum likelihood estimates.

```{r}
mean(x)
sd(x)*(n-1)/n
```

The log likelihood.

```{r}
mu <- seq(8,12,length=101)
sig <- seq(3,5,length=101)
ll <- matrix(NA,length(mu),length(sig))
for(i in 1:length(mu)){
  for(j in 1:length(sig)){
    ll[i,j] <- sum(dnorm(x,mu[i],sig[j],log=TRUE))
  }
}
```

Plot the log likelihood.

```{r}
par(las=1)
image(mu,sig,ll,col=rev(rainbow(start=0,end=2/3,250)),
      xlab="", ylab="", main="n=100")
mtext(side=1,expression(mu),line=3,cex=1.5)
mtext(side=2, expression(sigma),line=2.5,cex=1.5)
contour(mu,sig,ll,nlev=16,add=TRUE)
points(mean(x),sd(x)*sqrt(99/100),lwd=2,pch=4,cex=1.3,col="blue")
arrows(mean(x)+0.8, sd(x), mean(x)+0.2, sd(x), lwd=2,col="blue",len=0.15)
text(mean(x)+0.9,sd(x), "MLE", adj=c(0,0.5), col="blue", cex=1.4)
```

### Example from class: 2-point linkage in an intercross

The log likelihood function for the 2-point intercross data.

```{r}
llf2 <- function(dat,rf=seq(0,0.5,len=501)){
  rfbar <- 1 - rf
  A <- dat[1,1]+dat[3,3] # non-recombinants
  B <- dat[1,2]+dat[2,1]+dat[2,3]+dat[3,2] # single-recombinants
  C <- dat[1,3]+dat[3,1] # double-recombinants
  D <- dat[2,2] # double heterozygotes
  A * log(rfbar^2/4) + B * log(rf*rfbar/2) + C*log(rf^2/4) +
    D*log((rf^2+rfbar^2)/2)
}
```

The example from class.

```{r}
mydat <- rbind(c(58,9,0), c(8,95,14), c(1,12,53))
mydat
```

The function "optimize" is for finding the maximumor minimum of a function of one variable.

```{r}
mle <- optimize(llf2,c(1e-5,0.4),dat=mydat,maximum=TRUE)$maximum
mle
```

The log likelihood function for these data.

```{r}
rf <- seq(0,0.5,len=501)
ll <- llf2(mydat)
plot(rf,ll,type="l",lwd=2,xlab="Recombination Fraction",ylab="log likelihood")
abline(v=mle,col="blue",lwd=2)
u <- par("usr")
text(mle+diff(u[1:2])*0.1, mean(u[3:4]), "MLE = 9.4%", cex=1.5, col="blue")
```

A closer view of the log likelihood function.

```{r}
rf <- seq(0.05,0.15,len=501)
ll <- llf2(mydat,rf)
plot(rf,ll,type="l",lwd=2,xlab="Recombination Fraction",ylab="log likelihood")
abline(v=mle,col="blue",lwd=2)
u <- par("usr")
text(mle+diff(u[1:2])*0.1, mean(u[3:4]), "MLE = 9.4%", cex=1.5, col="blue")
```

