```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.align="center")
options(width=110)
```

#### Statistics for Laboratory Scientists ( 140.615 )

## Summarizing and presenting data

Data from Sokal & Rohlf, exercise 4.2 (pg 59) 
[ mg glycine per mg creatine in the urine of 37 chimpanzees ]

```{r}
x <- c(0.008, 0.018, 0.056, 0.055, 0.135,
       0.052, 0.077, 0.026, 0.044, 0.300,
       0.025, 0.036, 0.043, 0.100, 0.120,
       0.110, 0.100, 0.350, 0.100, 0.300,
       0.011, 0.060, 0.070, 0.050, 0.080,
       0.110, 0.110, 0.120, 0.133, 0.100,
       0.100, 0.155, 0.370, 0.019, 0.100,
       0.100, 0.116)
x
length(x)
str(x)
class(x)
```

#### Some summary statistics

```{r}
mean(x)
median(x)
min(x)
max(x)
range(x)
summary(x)
quantile(x)
quantile(x,probs=c(0.25,0.75))
IQR(x)
sd(x)
mad(x)
```

#### Some plots

##### Histograms with different numbers of bins

```{r}
hist(x)
hist(x, breaks=5)
hist(x, breaks=15)
hist(x, breaks=seq(0,0.4,0.05))
```

##### Plot values by index

```{r}
plot(x)
plot(x, type="l")
plot(x, type="b")
```

##### Dot plots

```{r}
stripchart(x)
set.seed(1)
stripchart(x, method="jitter")
stripchart(x, method="jitter", pch=1)
stripchart(x, method="jitter", pch=1, cex=1.5)
stripchart(x, method="jitter", pch=1, cex=1.5, log="x")
stripchart(x, method="jitter", pch=1, cex=1.5, log="y", vertical=TRUE)
```

You can change the orientation of the label of the y-axis with the par() graphics function.

```{r}
par(las=1)
stripchart(x, method="jitter", pch=1, cex=1.5, log="y", vertical=TRUE)
```

You can specify the width of the rendered plot in the code chunk start.

```{r, fig.width=3}
par(las=1)
stripchart(x, method="jitter", pch=1, cex=1.5, log="y", vertical=TRUE)
```

##### Boxplots

```{r}
boxplot(x)
boxplot(x, range=0)
boxplot(x, range=0, log="y")
boxplot(x, range=0, log="y", col="lightblue")
```

#### Arithmetic and geometric mean

```{r}
mean(x)
exp(mean(log(x)))
```

#### Subsetting and sorting

```{r}
x[1]
x[-1]
x[1:5]
x[-(1:5)]
x[c(1,7,22)]
x<0.1
x[x<0.1]
sort(x)
sort(x, decreasing=TRUE)
order(x)
order(x, decreasing=TRUE)
```



